---
title: "PML Project Writeup"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
author: 
date: "September 18, 2015"
---
### Executive Summary 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively inexpensively. 
These type of devices are part of the quantified self movement – a group of enthusiasts 
who take measurements about themselves regularly to improve their health, to find 
patterns in their behavior, or because they are tech geeks. One thing that people 
regularly do is quantify how much of a particular activity they do, but they rarely 
quantify how well they do it. 

In this project, the goal will be to use data from accelerometers on the belt, forearm, 
arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly 
and incorrectly in 5 different ways. More information is available from the website 
here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting 
Exercise Dataset).

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from the paper referenced below and downloaded here: http://groupware.les.inf.puc-rio.br/har . 

Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable 
Computing: Accelerometers' Data Classification of Body Postures and Movements. 
Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in 
Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. 
Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. 
DOI: 10.1007/978-3-642-34459-6_6.  


###Data Processing 
__Load required libraries__
```{r setup}
library(caret)
```   
The data for the project has been downloaded as csv files from the links refernced in the executive summary.    
__Load the training data__
```{r trn_load}
x <- read.csv("pml-training.csv", na.strings=c("NA", ""))
dim(x)
x[1:5,1:5]
```  
__Load the test data__
```{r tst_load}
y <- read.csv("pml-testing.csv", na.strings=c("NA", ""))
dim(y)
y[1:5,1:5]
```  
__Cleaning the data__ is next, and first we need to remove all the columns with NA in them because they are summary based.  
```{r clear_NA}
x <-x[,colSums(is.na(x))==0]
y <-y[,colSums(is.na(y))==0]
```  
Our next step is to convert to numeric data types and remove columns we know are not features. 
```{r num_nuke}
###convert all the features to numerics
for(i in 8:59){tmp <- typeof(x[,i]); if(tmp == "integer") x[,i]<- as.numeric(x[,i])}
for(i in 8:59){tmp <- typeof(y[,i]); if(tmp == "integer") y[,i]<- as.numeric(y[,i])}
#### remove un-needed columns
x<- x[,c(7:60)]
y<- y[,c(7:60)]
dim(x)
dim(y)
```

### Exploratory Data Analysis
```{r expDA}
dim(x)
names(x)
summary(x$classe)
plot(x$classe, col=rainbow(5), main="x$classe Frequency")
```

At this point we are ready to split the training set into two so we can have cross validation data later.  
```{r split}
### split into training and validation sets 
inTrain <- createDataPartition(x$classe, p=0.70, list=FALSE)
trainx<- x[inTrain,]
validx<-x[-inTrain,]
dim(trainx)
dim(validx)
```
### Model Fitting 
We plan to use random forest for the model fit because of it's high level of accuracy. We also with use the cross validation option with 5 folds. 
```{r fit, cache=TRUE}
### fit a model using random forest
fit <- train(classe~.,data=trainx, method="rf", trControl = trainControl(method="cv", number = 5), prox = TRUE)
fit
```

### Calculating Accuracy in/out of sample

```{r accuracy_in,cache=TRUE}
train_ans<-predict(fit,trainx)
confusionMatrix(train_ans, trainx$classe)
```
 The confusionMatrix above shows that our in_sample accuracy is equal to 1 which is 100%. 
 
```{r accuracy_out,cache=TRUE} 
valid_ans<-predict(fit,validx)
confusionMatrix(valid_ans, validx$classe)
``` 
As we would expect the confusionMatrix above shows our out of sample accuracy is slightly less at .9983 which is 99.83%. 

### Predicting the test dataset 
As required for the project we now will predict the 20 values of the test case.
```{r final_ans,cache=TRUE}
answer <- as.character(predict(fit,newdata=y))
answer
```   
### Write out the assignment files as required. 

```{r write_files,cache=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answer)

```